# Course 5: Introduction to R {-#intro-r}

One of the most important exercises for this course is to make students comfortable with RStudio and RStudio cloud. These are some of the exercises you can ask students to do during the class.

* Uploading files on RStudio.cloud
* Exporting (downloading) files on RStudio.cloud
* Renaming files in the Files section
* Deleting files in the Files section
* Creating new folders in the Files section
* Looking at the packages installed
* Looking at the environment through the Environment tab
* Looking at history through the History tab
* Browsing over the Console and the Terminal
* Creating a new R script file
* Creating a new R Markdown file
* Opening a new R script file


## Exercise 1: Installing and Uninstalling Packages {-}

<div class = "tip">
Level: Easy
</div>

Start with telling students what CRAN is. CRAN hosts additional packages that sit on top of the core (base) R software. While there are thousands of packages on CRAN, a lot of the packages are hosted on other hosting sites such as Github. In this exercise, we will ask students to install packages from both CRAN and Github.

First, ask students to install the **devtools** package. Ask them to use the package name without the quotations as below:

```{r eval=FALSE}
install.packages(devtools)
# Error in install.packages : object 'devtools' not found
```

They will get an error saying the "object 'devtools' not found". Remind them when we install packages, we have to put the package name in quotations. So the correct way is:

```{r eval=FALSE}
install.packages("devtools")
```

The package **Devtools** is hosted on CRAN, therefore, we don't have to specify where the package is hosted. We just have to pass the name of the package to the function `install.packages()` to install it. Now ask them to load the package. Remind them that to load packages we use the following function:

```{r}
suppressWarnings(library(devtools))
```

Note that when we load packages we don't necessarily have to use quotation marks. It is recommended, however, that to use quotations for both `install.packages()` and `library()` functions. The package **devtools** makes package development easier. One of the applications of the package is that it allows us to install packages that are not on CRAN. Now ask students to install the package **knockknockjokes** from github using. The link to the Github repo of the package is [www.github.com/psolymos/KnockKnockJokes](www.github.com/psolymos/KnockKnockJokes). Ask them how they are supposed to install a package from Github.

```{r}
suppressWarnings(devtools::install_github("psolymos/KnockKnockJokes"))
```

Note that we have to specify where the package is hosted when we use the command `install_github()` from the package **devtools**. psolymos is the username of the user who developed the package and KnockKnockJokes is the name of the repository. Now ask students to load the package. 

```{r}
suppressWarnings(library(KnockKnockJokes))
```

This package is very simple. It basically tells you knock knock jokes. You can ask students to browse through the package documentation on [its Github repo](https://github.com/psolymos/KnockKnockJokes) and see what function they should use for a joke. Here's one for a random joke:

```{r}
KnockKnock()
```

As an assignment (in class or at home), you can ask students to find an R package they find interesting and tell the class about it.

## Exercise 2: Using the script editor {-}

<div class = "tip">
Level: Easy
</div>

Ask students to go to the my_music_project on RStudio.cloud that they previously created. Ask them to create a new R script file in the folder code/tidy_code and save it as exploratory_analysis. Ask them to install and load the package **dplyr** in the beginning of the file. One of the nice things about RStudio is its keyboard shortcuts. One that is used a lot is Ctrl+Enter.

Tell your students to type in `2 + 2` in the script file and while the cursor is on that line, hold the Ctrl (Command) key and then press Enter (or Return). What happens? Ctrl+Enter is used for running code without clicking on the Run tab. It makes running codes line by line a lot easier. As long as the cursor is somewhere on the line, Ctrl+Enter will run that line. 

You can also remind them that they can select and highlight multiple line and use Ctrl+Enter. For instance, ask them to type these two lines in an R script file and highlight both line together and then use Ctrl+Enter.

```{r}
x <- 2
x + 3
```


## Exercise XX: Basic Commands in R {-}

<div class = "tip">
Level: Easy
</div>

In this exercise, we will perform some basic commands in R on simple vectors. First, ask your students to create an vector object called `ages` that contains the following numbers:

```{r}
ages <- c(22, 18, 20, 25, 22, 19, 32, 17)
```

After giving them the above numbers and saving the object, ask your students to use an R function to find whether there is anybody who is older than 30 years old. Tell them you simply just need a TRUE/FALSE answer to this quesion. They can use the following command:

```{r}
any(ages > 30)
```

Now, ask them to use an R function to find whether all ages are above 15. Again, remind them that the answer should be a simple TRUE/FALSE. They should use something like:


```{r}
all(ages > 15)
```

You can further ask them to find which ages are larger than 20 For this, they can use the function `which()` to obtain the index of the element that is larger than 20 like below.

```{r}
a <- which(ages > 20)
a
ages[a]
```



## Exercise XX: Creating a Data Frame {-}

<div class = "tip">
Level: Medium
</div>

In this exercise, students will learn how to create a data frame. But since they've already learned how to create a repository on Github and push things to Github, ask them to create a repositiory there called movie_project. Further, ask them to clone the repository to an RStudio.cloud project with the same title. In the RStudio project, they should create all the necessary folders as they've learned before.

We are going to create a data frame that has four columns: `title`, `year`, `rating`, and `genre` for the top 10 movies on imdb. Ask your students to go to the link [here](https://www.imdb.com/search/title?genres=drama&groups=top_250&sort=user_rating). Ask them to create an R script file in the folder called code and type their code there. Based on the information on the link, ask them to input the title, year, rating, and genre information for the top 10 movies in a data frame. Ask them to call the data frame `movies_df` They should do something like:

```{r}
title <- c("The Shawshank Redemption", "The Godfather", "The Dark Knight", "The Godfather: Part II",
                "The Lord of the Rings: The Return of the King", "Pulp Fiction", "Schindler's List",
                "12 Angry Men", "Fight Club", "The Lord of the Rings: The Fellowship of the Ring")
year <- c(1994, 1972, 2008, 1974, 2003, 1994, 1993, 1957, 1999, 2001)
rating <- c(9.3, 9.2, 9.0, 9.0, 8.9, 8.9, 8.9, 8.9, 8.8, 8.8)
genre <- c("Drama", "Crime", "Action", "Crime", "Adventure", "Crime", "Biography", "Drama", "Drama", "Adventure")
movies_df <- data.frame(title, year, rating, genre)
```

Now that they've created the data frame, ask them to show the first 4 rows of the data:

```{r}
head(movies_df, 4)
```

Ask them to call the column that contains information about movie titles. 

```{r}
movies_df$title
```

Ask them to find the "class" of the columns year and genre:

```{r}
class(movies_df$year)
class(movies_df$genre)
```

Ask your student to find the difference between the maximum and the minimum rating in the data.

```{r}
dif <- max(movies_df$rating) - min(movies_df$rating)
dif
```

Then ask them to find the dimension of the data frame:

```{r}
dim(movies_df)
```

Now ask them to change the column called `title` to `name`:

```{r}
colnames(movies_df)[1] <- "name"
```

Without checking manually, ask them to see if the movie “Groundhog Day” is among the top 10 movies:

```{r}
any(movies_df$name == "Groundhog Day")
```

Ask them to find the number of the movies are in the genre Drama:

```{r}
drama <- which(movies_df$genre == "Drama")
length(drama)
```

```{r}
d <- list(movies = c("m1", "m2"),
     states = 4)
```

## Exercise XX: Creating Lists in R {-}

<div class = "tip">
Level: Easy
</div>

The beauty of lists are that you can have elements with different sizes and structures in them. For this exercise, ask your students to create a list with three elements. The first element called `places` should be a vector of the states (provinces) or countries your students have been to. The next element called `ice_cream` should conain the name of their top two ice cream flavors. The last element called `sports` should contain their favorite sports. The list should be called `about_me`. The code should look like:

```{r}
about_me <- list(places = c("New Mexico", "Vermont", "Washington", "California"),
                 ice_creams = c("Vanilla", "Chocolate"),
                 sports = c("Soccer"))
```

Ask them to call the first element of the list:

```{r}
about_me[[1]]
## note that they can also call it by 
about_me[["places"]]
```

Now, ask then to add another sport that they like to the element `sports`.

```{r}
about_me[["sports"]][[2]] <- "Volleyball"
```

## Exercise XX: Missing Values in R {-}

<div class = "tip">
Level: Moderate
</div>

Ask your students to load the `airquality` dataset in R. Ask them to show the columns that exist in the data and find the dimension of the data.

```{r}
colnames(airquality)
dim(airquality)
```

Ask them to look at the first 10 rows of the data:

```{r}
head(airquality, 10)
```

Do they see any missing values in the first 10 rows? They should answer yes as the the first 10 rows clearly has some missing values. Now, ask them to use the appropriate function that checks whether the column `Temp` has missing values. How about the column `Ozone`?

```{r}
any(is.na(airquality$Temp))
any(is.na(airquality$Ozone))
```

The column Ozone, contains mean ozone in parts per billion from 1300 to 1500 hours at Roosevelt Island. What rows in the column `Ozone` have missing values

```{r}
which(is.na(airquality$Ozone))
```

Now, ask your students to use the appropriate function to calculate the average of the column `Ozone`.

```{r}
mean(airquality$Ozone)
```

If they use the code above, the answer will be `NA`. Tell them that this is due to the missing values in the column and because of that, the function `mean()` does not calculate the average. For this, to see the average, they will have to use the argument `na.rm` and set it to `TRUE`. By doing this, they will tell R to ignore the missing values and only perform the function on the non-missing values.

```{r}
mean(airquality$Ozone, na.rm = TRUE)
```

Now, ask them to remove all the rows containing missing values from the entire data and save the new object as `airquality_nomissing`. Note that they should use the function `na.omit()`. What is the dimension of the new data?

```{r}
airquality_nomissing <- na.omit(airquality)
dim(airquality_nomissing)
```

Finally, ask them to write code that will replace all values of the column `Wind` that are equal to `8.0` with `NA`.

```{r}
airquality$Wind[airquality$Wind == 8.0] <- NA
```

## Exercise 4: Reverse a String Function {-}

<div class = "tip">
Level: Difficult
</div>

Ask your students to write a function that takes a string and returns the reverse version of the string. So the function returns the string "book" as "koob". This function is a simple version of a decoding function. Let's decompose this problem. There are probably various ways to do this problem. One way to go about solving this problem is to decompose the words into characters. Then reverse the order of the characters and then paste them back together. The first step can be done using the `str_split()` function from the **stringr** package.

```{r}
library(stringr)
word <- "book"
word_decom <- str_split("book", "")
word_decom
```

The `word_decom` object is a list and its first element is a vector of all the character that make up the word *book*. If we have a vector, we can use the function `rev()` to reverse the order of its elements. So `rev(c(1,2,3))` will return `3 2 1`. So to reverse the order of characters in the object `word_decom` we write

```{r}
word_decom_rev <- rev(word_decom[[1]])
word_decom_rev
```

`[[1]]` is for calling the vector in the list. Now, we have the decomposed word reversed. But the result is a vector of characters and not a string. To create a string from all the characters, we can use the function `paste()` as we have used before. 

```{r}
word_rev <- paste(word_decom_rev, collapse = "")
word_rev
```

Your students now can combine all the steps and create the function. This would be the simplest form of the functino.

```{r}
word_reverse <- function(x){
    word_decom <- str_split(x, "")
    word_decom_rev <- rev(word_decom[[1]])
    word_rev <- paste(word_decom_rev, collapse = "")
    return(word_rev)
}
word_reverse("chromebook")
```


## Exercise 5: Other Short Exercises {-}

<div class = "tip">
Level: Easy to moderate
</div>


Write a function that takes user's name and returns "Hello, Name!"

```{r}
hello <- function(name){
    print(paste("Hello ", name, "!", sep = ""))
}
```

Write a function that takes number 1 and number 2 and checks whether number 1 is divisible by number 2. If divisible, returns a message saying the first number is divisible by the second number and if not returns a message saying the first number is not divisible by the second number.

```{r}
check_division <- function(num1, num2){
    if (num1 %% num2 == 0) {
        print(paste(num1, " is divisible by ", num2, ".", sep=""))
    } else {
        print(paste(num1, " is NOT divisible by ", num2, ".", sep = ""))
    }
}
```

In the function above, check to make sure number 1 is bigger than number 2. If not, halt and show a message that the first number should be larger than the second number.


```{r}
check_division <- function(num1, num2){
    if(num1 < num2) stop("The first number should be larger than the second number.")
    else {
       if (num1 %% num2 == 0) {
        print(paste(num1, " is divisible by ", num2, ".", sep=""))
    } else {
        print(paste(num1, " is NOT divisible by ", num2, ".", sep = ""))
    } 
    }
}
```





